---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r libraries, include=FALSE}
library(tidyverse)
```

Import the data using the read_csv function that is part of the tidyverse
package. In most database systems, lack of data (not zero, but specifically a
lack of data) are represented by `NULL`. In R, lack of data are represented by
NA. The `read_csv` package does not automatically convert strings to factors as is
the case for many other import methods, such as base R's `read.csv`.

```{r import-data, include=FALSE}

mcdowellArthropods <- read_csv(file = 'https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-cap.643.2&entityid=4bc08a24e06acfe761553cf8360b983d')
```

Inspect the data - what type of entity is this?

```{r check-class}

class(mcdowellArthropods)
```

Inspect the data - what is the structure?

```{r check-structure}

str(mcdowellArthropods)
```

Inspect the unique values. In this case, we are using the `apply` function to
"apply" the `unique` function over the margins of a data frame. The third argument
of the function (`unique`) indicates that we want the unique values of the
columns. That we want the unique values of the columns is indicated by the
second argument ("2") of this function; we would have used a 1 if we wanted the
unique values of each row (but that would not make sense in this case).

```{r unique-value}

apply(X = mcdowellArthropods, MARGIN = 2, FUN = unique)
```

Why are some trapcounts so low? We need to take a closer look at those. We will
use the filter function in dplyr. Here, we are passing the mcdowellArthropods
data object to the `filter` function, which is part of the dplyr package. dplyr
uses pipes (%>%) to link commands, a particularly nice feature when you are
performing many actions on an object.

```{r check-low-trapcount-dplyr}

mcdowellArthropods %>% 
  filter(trap_count < 6) %>%
  print(n = 50)
```

The base R version of performing the above (code below) is not onerous but
addressing a chunk like organisms-per-site-season (below) would be very verbose and complicated without
pipes and dplyr.

```{r check-low-trapcount-base}

print(mcdowellArthropods[mcdowellArthropods$trap_count < 6,], n = 50)
```

With a better sense for these data, we can modify them for our analyses.

First, we need to filter to the desired set of data. We want to (1) exclude collections with a low trap count as noted in the previous step, and (2) omit sites that were discontinued. We can identify sites that have been discontinued by looking at the most recent sampling date.

```{r check-low-trapcount-dplyr}

mcdowellArthropods %>% 
  group_by(site_code) %>%
  summarise(maxDate = max(sample_date)) %>% 
  arrange(maxDate)
```

Second, for our analyses, we can ignore the different size classes so all
organisms, regardless of size, can be summed for each sample.

Third, we want to bin the sampling dates by season and extract the year for
plotting.

Finally, we need to standardize the sampling effort to account for an
inconsistent number of traps across sampling events, here we will use the
average number of organisms per trap per, per season across the years of study.

```{r organisms-per-site-season}

organismsSiteCollection <- mcdowellArthropods %>% 
  filter(
    trap_count >= 6,
    !grepl('mine|prospector', site_code, ignore.case = TRUE)
  ) %>%
  rowwise() %>% 
  mutate(allsizes = sum(lt2mm, `_2_5mm`, `_5_10mm`, gt10mm, unsized, na.rm = T)) %>% 
  ungroup() %>%
  mutate(
    season = case_when(
      as.integer(format(sample_date, format = '%m')) %in% c(11,12,1,2) ~ 'winter',
      as.integer(format(sample_date, format = '%m')) %in% c(3,4) ~ 'spring',
      as.integer(format(sample_date, format = '%m')) %in% c(5,6,7,8) ~ 'summer',
      as.integer(format(sample_date, format = '%m')) %in% c(9,10) ~ 'fall'
    ),
    year = format(sample_date, format = '%Y')) %>%
  group_by(site_code, season, year) %>%
  summarise(average = sum(allsizes / trap_count))
```

We will use ggplot to plot our data, here plotting each site in its own panel.

```{r plot-by-seaons}

ggplot(organismsSiteCollection, aes(x=year, y=average, fill = factor(season))) + 
  geom_bar(stat = 'identity', position = position_dodge2(preserve = 'single')) + 
  facet_wrap(~site_code, nrow = 4, ncol = 2) +
  labs(x = "year") +
  labs(y = "average number of organisms per trap per season") +
  labs(fill = 'season') +
  ggtitle("McDowell Sonoran Preserve ground-dwelling arthropods: 2012-2019") +
  # scale_fill_manual(values = c('#e6e6e6', '#cccccc', '#999999', '#000080')) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```
