---
title: "Notebook: analysis of arthropods in the McDowell Sonoran Preserve"
data: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
library(readr)
library(skimr)
library(ggplot2)
library(dplyr)
knitr::read_chunk("mcdowell_arthropods_functions.R")
```

## data import

Import the data from the Environmental Data Initiative data catalog using the
read_csv function that is part of the tidyverse package. In most database
systems, lack of data (not zero but specifically a lack of data) are represented
by `NULL`. In R, lack of data are represented by NA. The `read_csv` package does
not automatically convert strings to factors as is the case for many other
import methods, such as base R's `read.csv` (prior to R version 4.0.0).

```{r import-data, include=FALSE}
mcdowellArthropods <- readr::read_csv(
  file = 'https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-cap.643.2&entityid=4bc08a24e06acfe761553cf8360b983d'
)
```

## inspect the data

Inspect the data: what type of entity is this?

```{r check-class}
class(mcdowellArthropods)
```

Inspect the data: we can use the `str()` function to get a sense of the data
structure (in the Tidyverse, `dplyr::glimpse()` is a nice option for this as
well).

```{r check-structure}
str(mcdowellArthropods)
```

`str()` is great but limited. We can use the `skimr` package to quickly get a
more comprehensive look into the data.

```{r unique-value}
skimr::skim(mcdowellArthropods)
```

## explore low trap counts

Why are some trap counts low? We need to take a closer look at those. We will
use the filter function in dplyr. Here, we are passing the mcdowellArthropods
data object to the `filter` function, which is part of the dplyr package. dplyr
uses pipes (%>%) to link commands, a particularly nice feature when you are
performing many actions on an object.

We can include this chunk in the Rmarkdown document sensu below -- or --

```{r check-low-trapcount-within-doc, eval=FALSE}
mcdowellArthropods %>% 
  filter(trap_count < 6) %>%
  select(site_code:trap_sampling_events_comments, -trap_name, -observer) %>%
  print(n = 50)
```

we can reference an external function sensu below.

```{r, check-low-trapcount}
```

*Note: the base R version of performing something simple like the above is not onerous but the Tidyverse work flow makes addressing a more complicated chunk like organisms-per-site-season (below) much easier.*

## modify data as needed for analysis

With a better sense for these data, we can modify them for our analyses.

First, we need to filter to the desired set of data. We want to (1) exclude
collections with a low trap count as noted in the previous step, and (2) omit
sites that were discontinued. We can identify sites that have been discontinued
by looking at the most recent sampling date.

```{r check-last-sample-collection}
mcdowellArthropods %>% 
  group_by(site_code) %>%
  summarise(maxDate = max(sample_date)) %>% 
  arrange(maxDate)
```

Second, for our analyses, we can ignore the different size classes so all
organisms, regardless of size, can be summed for each sample.

Third, we want to bin the sampling dates by season and extract the year for
plotting.

Finally, we need to standardize the sampling effort to account for an
inconsistent number of traps across sampling events, here we will use the
average number of organisms per trap per, per season across the years of study.

```{r organisms-per-site-season}
organismsSiteCollection <- mcdowellArthropods %>% 
  filter(
    trap_count >= 6,
    !grepl('mine|prospector', site_code, ignore.case = TRUE)
  ) %>%
  rowwise() %>% 
  mutate(allsizes = sum(lt2mm, `_2_5mm`, `_5_10mm`, gt10mm, unsized, na.rm = T)) %>% 
  ungroup() %>%
  mutate(
    season = case_when(
      as.integer(format(sample_date, format = '%m')) %in% c(11,12,1,2) ~ 'winter',
      as.integer(format(sample_date, format = '%m')) %in% c(3,4) ~ 'spring',
      as.integer(format(sample_date, format = '%m')) %in% c(5,6,7,8) ~ 'summer',
      as.integer(format(sample_date, format = '%m')) %in% c(9,10) ~ 'fall'
    ),
    year = format(sample_date, format = '%Y')) %>%
  group_by(site_code, season, year) %>%
  summarise(
    average = sum(allsizes / trap_count),
    .groups = 'drop'
  )
```

## plot the data

We will use ggplot to plot our data, here plotting each site in its own panel.

```{r plot-by-seaons}

ggplot(organismsSiteCollection, aes(x=year, y=average, fill = factor(season))) + 
  geom_bar(stat = 'identity', position = position_dodge2(preserve = 'single')) + 
  facet_wrap(~site_code, nrow = 4, ncol = 2) +
  labs(x = "year") +
  labs(y = "average number of organisms per trap per season") +
  labs(fill = 'season') +
  ggtitle("McDowell Sonoran Preserve ground-dwelling arthropods: 2012-2019") +
  # scale_fill_manual(values = c('#e6e6e6', '#cccccc', '#999999', '#000080')) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```
